Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-10 analysis/paper/paper.Rmd
edit GM section

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes (Adams et al., 2004; Bookstein, 1997; Lawing and Polly, 2010; Slice, 2007). Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches (Adams et al., 2004). Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features (Birch and Martinón-Torres, 2019; Buchanan et al., 2019; Cardillo, 2010; Haruda et al., 2019; Lycett and Cramon-Taubadel, 2013; Meloro et al., 2015).

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals (Bookstein, 1997; Lawing and Polly, 2010). Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy (Bookstein, 1997; Gunz and Mitteroecker, 2013; Slice, 2007). Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients (Kuhl and Giardina, 1982). EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions (Adams et al., 2004; Bonhomme et al., 2014; Claude, 2008). The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics (Bonhomme et al., 2014). EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics (Fox, 2015; Hoggard et al., 2019; Ioviţă, 2010; Topi et al. 2018; Selden Jr 2019; Wilczek et al. 2014). Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.

>>>>

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or visually distinctive osteological features [@Birch2019; @Buchanan2019; @Cardillo2010; @Haruda2019; @Lycett2013; @Meloro2015].

However, a key limitation of the landmark approach is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approach is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Gunz2013; @Slice2007]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology, stone artifacts, and ceramics [@Fox2015; @Hoggard2019; @Iovictua2010; @Topi2018; @Selden2019; @Wilczek2014]. Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standardization from different periods.
